Static contract checking for Haskell

### 2011-10-21 paper.tex
Paper: done editing section 3, and mostly done overall.
Added extensive notes about min, including the beginnings of an alternative translation
that seems more natural to me.

Consider:\note[nc]{Unlike the POPL '09 paper, $CF(e)$ is not implied by $e \in \{x|p\}$.}
\begin{verbatim}
data List a = Nil | Cons a (List a)

@@

\begin{itemize}
\item $(\Phi_1)$ For each $K_i$ of arity $a_i$ we introduce selectors
  $sel_{k,K_i}$, which are the projection of $K_i(x_1,\dots,x_{n_i})$
  on its $k$-th component, $k = 1\text{ to }a_i$.
\item $(\Phi_2)$ For each pair of constructors $K_i,K_j$, $i \not = j$, we state that they can
  never map to the same value.
\item $(\Phi_3)$ Then, we have to give crash-freeness conditions for each $K_i$:
  Notice that we have a equivalence.
  \begin{itemize}
  \item $\leftarrow$: if we pack crash-free values in a data
    constructor, the resulting value is crash-free.
  \item $\rightarrow$: a value $t$ of type $T$\note[nc]{
    ``type'' is used intuitively here: it means the constructor head is of type $T$.
    } is crash-free implies
    that every value packed in it is crash-free. Recall that one can
    define projection on any argument of a value of type $t$. So if
    the $k$-th argument of $t$ is not crash-free, then the $k$-th
    projection is a crash-free context that throws an expression that
    is not crash-free.

@@

\item $(\Phi_4)$ Now full application of a constructor $K_i$ is unreachable or crashing.
\item One may want to also state that if for all $j$, $x_j \neq \bad$ then
  $K_i(\vec{x}) \neq \bad$. It is already implied by the fact that
  $\left(\forall j. \cf{x_j}\right) \to \cf{K_i(\vec{x})}$.
\end{itemize}

@@

\subsection{$\ctrans{}$ -- Contracts}
We give in figure~\ref{ctrans} the translation of contract satisfaction.
$true$ refers to the translation to a term of the data constructor
\texttt{True} in $\hprime$, not to the actual true value.

Note that we define the translation of $f \in c$ and of $f \not \in
c$. We have to do that because one is not the negation of the other,
even though $\lnot (f \not \in c)$ implies $f \in c$.

>>>>

Consider:\note[nc]{Unlike the POPL '09 paper, $\cf{e}$ is not implied by $e \in \{x|p\}$.}
\begin{verbatim}
data List a = Nil | Cons a (List a)

@@

\begin{itemize}
\item $(\Phi_1)$ For each $K_i$ of arity $a_i$ we introduce selectors
  $sel_{k,K_i}$, which are the projection of $K_i(x_1,\dots,x_{a_i})$
  on its $k$-th component, $k = 1\text{ to }a_i$.
\item $(\Phi_2)$ For each pair of constructors $K_i,K_j$, $i \not = j$, we state that they can
  never map to the same value.
\item $(\Phi_3)$ Then, we have to give crash-freeness conditions for each $K_i$:
  Notice that we have a equivalence.
  \begin{itemize}
  \item $\leftarrow$: if we pack crash-free values in a data
    constructor, the resulting value is crash-free.
  \item $\rightarrow$: a value $t$ of type $T$\note[nc]{
    ``type'' is used intuitively here: it means the constructor head is of type $T$.
    } is crash-free implies
    that every value packed in it is crash-free. Recall that one can
    define projection on any argument of a value of type $t$. So if
    the $k$-th argument of $t$ is not crash-free, then the $k$-th
    projection is a crash-free context that throws an expression that
    is not crash-free.

@@

\item $(\Phi_4)$ No full application of a constructor $K_i$ is unreachable or crashing.
\item One may want to also state that if for all $j$, $x_j \neq \bad$ then
  $K_i(\vec{x}) \neq \bad$. It is already implied by the fact that
  $\left(\forall j. \cf{x_j}\right) \to \cf{K_i(\vec{x})}$.
\end{itemize}

@@

\subsection{$\ctrans{}$ -- Contracts}
We give in figure~\ref{ctrans} the translation of contract satisfaction.
\annote[nc]{$true$ refers to the translation to a term of the data constructor
\texttt{True} in $\hprime$, not to the actual true value.}{
Again, we should use $True$ (or \texttt{True}) in the translation; the
lowercase is an artifact of the concrete TPTP syntax.
}
\remove[nc]{Note that we define the translation of $f \in c$ and of $f \not \in
c$. We have to do that because one is not the negation of the other,
even though $\lnot (f \not \in c)$ implies $f \in c$.}\note[nc]{

The contract translation, $\ctrans{}$, introduces a new predicate,
$\min{}$, and contract non-satisfaction, $f \notin c$.  The formula
$\min{e}$, short for ``(a)m in(terested) in $e$'' (?), is used to
guide the proof search, by restricting proofs to follow the evaluation
order when instantiating hypotheses (Is this meaningful? Is it even
correct?).  To understand the translation intuitively, simply ignore
all $\min{}$'s (I.e., assume $\forall x. \min{x}$).  The
non-satisfaction form, $f \notin c$, is introduced to make the
$\min{}$'s appear positively in translations in negative/contravariant
positions; Ignoring $\min{}$'s, we have $\ctrans{f \notin c} \equiv
\lnot \ctrans{f \in c}.$ (Well, after I change the definitions of
$\ctrans{}$ for $e \in \{x | p\}$ and $e \notin \{x|p\}$ \ldots)

Dimitrios gave an example where defining $\ctrans{f \notin c}$ as
$\lnot \ctrans{f \in c}$ is problematic: prove $\ctrans{g ~Z \in
  \cfc}$, assuming $\min{g ~Z}$, for an abstract function $g \in \cfc
\to \cfc$.  If we define 
\protect \[ 
\ctrans{g \in x:c1 \to c2} :=
\forall x. \min{g x} \to \ctrans{x \in c1} \to \ctrans{g x \in c2},
\]
then we get stuck, because we must prove $\min{Z}$ to satisfy
$\ctrans{x \in c1}$ (i.e. $\min{Z} \land \cf{Z}$; we have $\cf{Z}$ by
definition).  On the other hand, if we define
\protect \[
\ctrans{g \in x:c1 \to c2} 
:= \forall x. \min{g x} \to \lnot \ctrans{x \notin c1} \to \ctrans{g x \in c2},
\]
then we succeed, because $\min{Z}$ is irrelevant to us in satisfying
$\ctrans{x \in c1}$ (i.e. $\lnot \min{Z} \lor \cf{Z}$).

The irrelevance of $\lnot \min{Z}$ above leads to the question, can we
do away with the $\min{}$'s altogether in such positions?  It turns
out the answer is ``No'', when we consider proving $g ~(g ~Z) \in
\cfc$, but I have some ideas for an alternative translation where the
answer is ``Yes''.  But first $g ~(g ~Z)$: to use $g$'s contract
assumption we need to prove $\lnot \min{g ~Z} \lor \cf{g ~Z}$, and to
prove $\cf{g ~Z}$ we in turn need to use $g$'s contract again.  To use
$g$'s contract again we need $\min{g ~Z}$, and so $\min{}$ is relevant
again.  The solution is to invoke excluded middle on $\min{g ~Z}$:
either $\min{g ~Z}$, and we're done, or $\lnot \min{g ~Z}$, in which
case we satisfy $g$'s preconditions \emph{without} showing $g$'s
argument to be crash free.  This seems a little surprising to me,
making a translation with no $\min{}$ here even more appealing.

Here's a start at such a translation.  I
think it's more natural to think about variance, so write
$\ctrans{}^+$ for conclusions and $\ctrans{}^-$ for assumptions.  Then
a system that works for these $\cfc$ examples is
\protect \begin{eqnarray*}
\ctrans{x \in \cfc}^+ &:=& \min{x} \to \cf{x} \\
\ctrans{x \in \cfc}^- &:=& \cf{x}\\
\ctrans{f \in x:c1 \to c2}^+ &:=& \forall x.               
  \ctrans{x \in c1}^- \to \ctrans{f x \in c2}^+\\
\ctrans{f \in x:c1 \to c2}^- &:=& \forall x. \min{f x} \to 
  \ctrans{x \in c1}^+ \to \ctrans{f x \in c2}^-
\end{eqnarray*}
Because $\Gamma,\phi \vdash \psi \equiv \Gamma \vdash \phi \to \psi$,
it's natural to think of contract assumptions in the theory as
negative, and contract goals as positive.  An example is helpful:
\protect \begin{eqnarray*}
\ctrans{g \in \cfc \to \cfc}^+ &=& \forall x. \cf{x} \to (\min{g ~x} \to \cf{g ~x}),\\
&&\text{whereas}\\
\ctrans{g \in \cfc \to \cfc}^- &=& \forall x. \min{g ~x} \to (\min{x} \to \cf{x}) \to \cf{g ~x}.
\end{eqnarray*}
Now, given $\ctrans{g \in \cfc \to \cfc}$, we may 
want to prove $\ctrans{g ~(g ~Z) \in \cfc}$ 
from $\ctrans{g \in \cfc \to \cfc}^-$ (note the negative).
So, our goal is $\min{g ~(g ~Z)} \to \cf{g ~(g ~Z)}$.
This time the proof proceeds more naturally (no EM).  
From $g$'s contract, we get $\cf{g ~(g ~Z)}$ if we can prove 
$\min{g ~Z} \to \cf{g ~Z}$.
But that follows by instantiating $g$'s contract again, and 
then using $\cf{Z}$ to get $\min{Z} \to \cf{Z}$.

Notice the assymetry between the positive and negative arrow
translations: the positive case has no $\min{f ~x}$ assumption.  I
left this $\min{}$ out because it's already in the 
conclusion of the $\cf$ example, but I certainly haven't
worked enough examples to justify it.  Moreover, this fragment
(no $e \in \{x|p\}$ yet) already fails to have a standard
property of bidirectional systems: that you can check/verify what you assume/use.
In other words, we can't prove that, forall $e$ and $c$,
\protect \[
  \ctrans{e \in c}^- \text{ implies } \ctrans{e \in c}^+.
\]
The trouble seems to come down to needing to prove $\min{e ~x} \implies
\min{x}$, which is not true a priori.  But, it's easy to add the $\min{g ~x}$ 
assumption to the positive arrow translation to recover the
 ``negative implies positive'' property.

Going back to the current translation, we consider some of it's
metatheoretic properties.
\protect \begin{enumerate}
\item $\lnot \ctrans{e \notin c}$ should behave like a double-negation
  translation, in the sense that $\ctrans{e \in c}$ should imply
  $\lnot \ctrans{e \notin c}$.  This becomes true if we modify the
  predicate case so that the implication holds (e.g. by more accurately
  modeling the operation semantics of contracts), and add a symmetric 
  $\min{f ~x}$ assumption to $\ctrans{f \notin x:c_1 \to c_2}$.  Note that
  reverse implication does not hold, e.g. $\lnot \ctrans{x \in \cfc}$ does
  not imply $\ctrans{x \in \cfc}$.
\item The ``fancy'' $\min{}$-ified translation should 
  degenerate to the ``naive'' translation when we assume $\forall x. \min{x}$.
  This is trivially true if we define the ``naive'' translatino as the ``min'' tranlation
  plus that axiom.  In symbols, we have
  \protect \[
  \Gamma^\text{min} \vdash \phi \implies \Gamma^\text{naive} \vdash \phi.
  \]  I.e., ``\min{} is sound for `naive'''.
\item We'd like to have a result  like
  \protect \[
  \text{Given some validity or consistency assumptions on $\Gamma$ and $e \in c$}, 
    \Gamma^\text{naive} \vdash \ctrans{e \in c}^\text{naive} \implies \Gamma^\text{min} \vdash \ctrans{e \in c}^\text{min}.
  \]  I.e., ``\min{} is complete for `naive', under some assumptions about 
  what we're proving''.  However, Dimitrios points out that we shouldn't expect this
  completeness theorem without some work, since we're purposefully 
  restricting the proof search to follow
  evaulation strategy, and it's far from obvious that all proofs can proceed this way.

\end{enumerate}

Concerns about the current translation
\protect \begin{enumerate}
\item The changes needed to make the first property above hold (a $\min{}$ in the negative
  arrow case and symmetric predicate contract cases).
  I had trouble proving
  \protect \[
  (\lambda x. \text{case $x$ of $True \to True$; $False \to \bad$}) 
  \in \{x|x==True\} \to \cfc
  \]
  , after adding $p = \unr$ to $\ctrans{x \in \{x|p\}}$, but I need to think about it more.

\item There may need to be some $\min{}$ assumptions in the prelude.  E.g. right now 
  it's impossble to prove $\ctrans{True \in \cfc}$, because you can't prove 
  $\min{True}$ a priori.  On the other hand, we don't need any extra 
  $\min{}$ assumption in the arrow case, since it's already assumed, at least in
  very simple examples (checked on the simple
  example that $id \in \cfc \to \cfc$).

\item A related question, is what do we feed to Equinox?  Equinox is a 
  refutation prover, so to prove $\Gamma \vdash \phi$ we ask if 
  $\Gamma,\lnot \phi \vdash \bot$.  So, for us, when $\phi$ is $\ctrans{e \in c}$, do
  give Equinox $\Gamma,\lnot \ctrans{e \in c}$ or $\Gamma, \ctrans{e \notin c}$.  I think
  we want the former, since ignoring Equinox our goal is $\Gamma \vdash \ctrans{e \in c}$.
\end{enumerate}
}
